\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity in Human Brains}

### 2014-01-16 docs/paper/paper final.tex
Added part of brain connectivity types to introduction.

\subsection{Previous research}
%[Tom: we did not explicitly have this paragraph, so this is additional effort]
%[Ramon]: Lijkt me vergelijkbaar met de verwijzingen naar literatuur die we al hebben? Met wat extra referenties naar wat Max zei: temporele methodes zijn onbetrouwbaar, dat hebben wij niet
%[Ramon]: even een goede balans zoeken tussen dit en "Existing problems" in motivation

@@

\section{Background}
\subsection{Types of brain connectivity}% [optional] [Ramon: lijkt me wel belangijk, geeft ook het belang aan van richting zoeken, hoeft niet lang]
%Functional
%Structural
%Effective

\subsection{Causal discovery theory}
%Causal Markov Condition
%Causal faithfulness
%d-separation
%V-structure
%Causal sufficiency
%Latent variables
An approach for finding analysing effective connectivity may be found in the domain of causal discovery.
A causal discovery method uses data of observed events to determine which of these events are correlated and additionally attempts to determine which events cause which.
We can define a causal relation as follows.

>>>>

\subsection{Previous research}
%[Tom: we did not explicitly have this paragraph, so this is additional effort]
%[Ramon]: Lijkt me vergelijkbaar met de verwijzingen naar literatuur die we al hebben? Met wat extra referenties naar wat Max zei: temporele methodes zijn onbetrouwbaar, dat hebben wij niet
%[Ramon]: even een goede balans zoeken tussen dit en "Existing problems" in motivation
% Previous work
Previous work on effective connectivity include the aforementioned Granger causality \cite{roebroeck2005}, Structural Equation Modelling \cite{mclntosh1994}, Multivariate Autoregressive Modelling \cite{harrison2003} and Dynamic Causal Modelling \cite{friston2003}.
All of these methods fail to cope with the problems mentioned in the introduction.
Most of these methods can only handle tens of variables in reasonable time, which becomes problematic in realistic situations with up to tens of thousands of variables.

@@

\section{Background}
\subsection{Types of brain connectivity}
Work on brain connectivity theory discerns three types of brain connectivity.

\paragraph{Functional connectivity}
Functional connectivity describes the statistical dependence of neuronal activity between different brain regions \cite{friston1993functional}.
Functional connectivity can be deduced through correlation of Functional Magnetic Resonance Imaging (fMRI) measurements of oxygen saturation  in blood - the BOLD response - while performing a certain cognitive task, or in so called resting-state.

\paragraph{Resting-state}
Resting-state indicates that subjects are instructed to relax without thinking of anything in particular to stimulate spontaneous brain activity; neural activity does not stop when not doing anything in particular.
Resting-state provides a typical whole brain behaviour with a low frequency ( < .1 Hz) that is reliable, robust, unique to a subject and task-independent.
Measuring resting-state thus provides an advantage when analysing brain connectivity on a global, whole brain level.
Resting-state measurements have the disadvantage that it is difficult to check whether a subject truly is in a resting-state; it is often quite difficult not to think of anything in particular.
Resting-state analysis has been shown to work well in practice \cite{Lowe2000, doria2010, Bullmore2009}.

\paragraph{Structural connectivity}
Structural connectivity is defined as the mapping of white-matter tracts in the brain between different regions \cite{friston1994}.
It is strongly related with functional connectivity: regions can only be functionally connected if there is a structural relation between them \cite{cabral2012}.
This intuitive relation can also be demonstrated empirically \cite{vandenheuvel2009}.
MRI scanners can be used to measure white-matter tracts directly.
As water is most concentrated in these regions, these tracts can be detected by measuring the diffusion of water in the brain.
This process is named Diffusion Weighted MRI (DWI).
For all practical purposes structural connectivity is time-independent - it changes over years, and still marginally during adulthood.

\paragraph{Effective connectivity}
In contrast to functional and structural connectivity, effective connectivity takes into account the cause and effect of connectivity relations.
It has been described as ``the influence one neural system exerts over another'' \cite{friston1994}.
Functional time series data needs to be analysed to infer effective connectivity as cause and effect can be deduced from which event precedes which.
Granger causality is a method relying on this temporal principle.
In brain data, this principle might not necessarily hold due to the fluctuations in haemodynamic response, the lag between neuronal activity and the measured BOLD response.
Nevertheless, in theory it is possible to infer effective connectivity from structural connectivity and functional data \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.
An established relation of effective connectivity to actual anatomical brain relations has yet to be established.

\subsection{Causal discovery theory}
An approach for finding analysing effective connectivity may be found in the domain of causal discovery.
A causal discovery method uses data of observed events to determine which of these events are correlated and additionally attempts to determine which events cause which.
We can define a causal relation as follows.
